Stack Safety is a Security Property

### 2021-04-07 tex/main.tex
Update based on some low-hanging-fruit comments

%% Title information
\title{Stack Safety as a Security Property}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
%                                        %% can be repeated if necessary;
%                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
%                                        %% can be repeated if necessary;
%                                        %% contents suppressed with 'anonymous'

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
\bcp{How about ``Stack Safety {\em is} a Security Property''?}
  What is stack safety? The phrase has been used to describe the protections
  offered by a variety of mechanisms for guarding against
  attacks that corrupt the call stacks of running programs, including micro-policies
  implemented using metadata tagging mechanisms such as PIPE. But for specification
  these implementations largely appeal either to intuitive notions of the behavior
  of a stack, or to specific examples of bad behavior.
  %
  We propose a formal definition phrased in the technical terminology of language-based
  security---as a combination of an integrity property (a caller's stack frame is
  invariant over calls) and a confidentiality property (callees lack knowledge of the
  contexts of their calls). These combine with a straightforward well-bracketedness
  condition on the program's control flow that ensures that callers and callees remain
  meaningfully distinct.
  %
  While our formalism is informed by the PIPE implementation, the primary properties reflect
  general principles of stack safety. We define additional properties tailored for specific
  contexts: a stronger property amenable to efficient testing and a weaker one that is
  approximated by more efficient, lazy policies.

@@

But what {\em is} the stack safety that these techniques enforce? Precise formal
definitions are mostly lacking: rather, the notion is commonly defined in a
negative way, by enumerating the bad things that a given enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%\bcp{Does use of unitialized variables lead to stack safety attacks?\apt{sure, via buffer overflow}}
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
Indeed, we are only aware of one quite recent attempt to frame a positive definition, by
\citet{Skorstengaard+19}, which defines stack safety as the conjunction of two properties:
{\em local state encapsulation} and {\em well-bracketed control flow}.
These properties are formalized by defining a capability-based machine
that ``satisfies [stack safety] by construction.''
%
While this work represents an important step in the right direction, we believe
there is still room to craft an operational characterization
of stack safety. \apt{Perhaps our rhetoric should be more micro-policy-specific here?}
In other words, we seek formal criteria that can be applied
to a mechanism for protecting stack-based low-level programs to judge
whether or not it actually ``guarantees stack safety.'' We begin with a strong
local state encapsulation property that closely models the behavior of eager
micro-policies from \citet{DBLP:conf/sp/RoesslerD18}. This ``eager'' property
is too strong to characterize stack safety; reasonable enforcement mechanisms
exist that violate it. One such mechanism is a ``lazy'' tag policy, similar to
the one in \citet{DBLP:conf/sp/RoesslerD18}. We weaken local state encapsulation
to support this technique and argue that the new property, defined relative to an
arbitrary notion of observation, is weak enough to characterize stack safety.
\apt{Reads oddly. ``string enough'' is what we want, surely? (While still weak
  enough to be enforced by the lazy policies.)}

While our criteria are designed with PIPE micro-policies in mind, they are
agnostic about many implementation details and should be applicable to other
systems with minor adjustment.
%
Our key contribution is an operational model of local state encapsulation
using the technical framework of language-based security\apt{this idea deserves more prominence}, with eager and lazy variants,
together with a simple characterization of well-bracketed control flow.

@@

How does the programmer rely on stack safety? The stack makes implicit promises about
the behavior of function calls. First, that when a call is finished, the caller can continue
from where it left off with its local state unchanged. And second, that the call is
self-contained, its behavior determined only by a predictable interface -- at minimum
parameters and global variables, but never local state.\apt{This last sentence is not so clear.}

@@

Both expectations imply predictable control flow, termed {\em well-bracketed
control flow} (WBCF) by \citeauthor{Skorstengaard+19}. Calls jump to a consistent \apt{??} entry
point and return to the call site. Then the data portion of the first expectation is for
{\em integrity}: that a caller's local data is protected from modification by its callee.
This is vital for local reasoning about the behavior of the caller. The security implications
are significant: without integrity, a caller that branches on a local variable could be manipulated by its
callee into unexpected behavior.

Alongside integrity, we have {\em confidentiality}, in which a caller's data is protected
from leakage by its callee. As a reasoning principle, confidentiality tells us that to
predict the behavior of a call we need only consider a subset of the machine state.
Rearranging local data will have no impact.\apt{??} And in security terms, we can guarantee
the secrecy of local data that we haven't passed even when calling a badly behaved function.
Together, integrity and confidentiality comprise \citeauthor{Skorstengaard+19}'s
{\em local state encapsulation} (LSE) property.

@@

To phrase LSE and WBCF as security properties requires an explicit attacker model:
a clear understanding of who is being protected and from what. Most existing
stack protection mechanisms envisage scenarios where a callee is
influenced by external inputs to corrupt the stack in some way. To
avoid getting into the details of exactly how this happens when
formalizing our properties, we use an even stronger model: when we
state the protections that a function can rely on when it makes a
call, we assume that the callee can execute arbitrary instructions
until it returns. Thus we have no need to differentiate safe code
from adversarial - all functions, and indeed all activations of the same function,
are mutually distrusting.
%
Even more fundamentally, \apt{``For this model to make sense,'' (??)} we need to know when a call or a return is taking
place. We assume that calls and returns are annotated in the code, so that
we can track the point at which control transfers from the protected caller
to the attacking callee -- and when that callee in turn becomes a protected
caller in its own right.

@@

First we introduce this model for a single simple stack with no sharing between callers and
callee. Then we expand it to allow components\apt{??} to be passed, granting a callee one time
access to the caller's data. We do not model more sophisticated sharing, because to do
so realistically would require a separate protection scheme, such as a capability
model. While this is quite feasible in a tag-based enforcement mechanism, it is properly
a separate policy outside the scope of stack safety. (The combination of such policies is
a strength of tag-based approaches.)

Finally, we extend the model to a coroutine system featuring finitely-bounded stacks. Though
simple\apt{simplistic}, this extension is a step toward concurrency and demonstrates that the model is flexible
enough to handle more sophisticated styles of control-flow.

After formally defining stack safety, we show that the definition is
enforceable using an existing tag-based enforcement
mechanism~\citep{DBLP:conf/sp/RoesslerD18}. We first consider
\citeauthor{DBLP:conf/sp/RoesslerD18}'s conservative Depth Isolation policy.
This policy testably enforces our {\em eager} properties, in which integrity is respected
at every step of a program, and each call respects integrity at each of its internal steps.
These properties are very strong, and our policies enforce them; they also have the advantage
that they ``fail faster'' \apt{than what?} and are therefore suitable for efficient testing.

The downside of the Depth Isolation policy is that it is rather expensive to implement.
In the same work, \citeauthor{DBLP:conf/sp/RoesslerD18} propose a different mechanism
that is more efficient by virtue of being lazy in its enforcement.
Instead of enforcing the integrity and confidentiality of the stack at the
exact moment when a callee returns, this policy signals violations when the
caller actually accesses data that the callee has written into its stack
frame. \ifaftersubmission\bcp{We could also hint at the lazy confidentiality
  property in the same way...}\fi
The lazy policy does not enforce our eager integrity property, and to our
knowledge there is no formal statement of the properties that it ought to enforce.
We propose {\em lazy} stack-safety properties, which capture the desired behavior of a
policy that detects integrity violations between when they occur and when they make
a visible impact. Intuitively this can be phrased as ``no visible reads after dangerous
writes.'' Unfortunately \citeauthor{DBLP:conf/sp/RoesslerD18}'s existing
``Lazy Tagging and Clearing'' policy does not quite enforce this property, as we show in
testing, but formalizing the ideal property allows us to see how it falls short.
We provide a variant of ``Lazy Tagging and Clearing'' that does enforce lazy integrity
at some performance cost. We prove (in Coq) that the eager property implies the lazy one.\apt{do we?
  and is it so interesting anyhow?}
We argue that any system purporting to enforce stack safety ought to obey \apt{``at least'' ?} xslazy stack safety.

@@

In this very strong attacker model, there is little obvious distinction between
the attacker and the defender. They are separated only by the control transfer
inherent in their calls and returns. Our stack-safety concepts therefore depend
critically on the notions of ``call'' and ``return,'' which are not fully explicit
in machine code (the same instruction opcode may be used in some places as part of
a call or return sequence and in other places for other purposes).
%
We assume we are given a machine language program annotated with the locations of
instructions representing calls, returns, and function entry points. Although this
information would typically come from a compiler together with the machine code,
we do not assume anything about their provenance. We do treat the execution of an
annotated instruction as in some sense intentional on the part of the current
function. No accidental calls or returns, and attackers control their own returns
but cannot change the annotations on the caller's code.\apt{cryptic}

@@

Alternatively, suppose an annotated instruction is not really an actual \apt{``call or'' ?} return.
What happens in this case depends on the enforcement mechanism and the nature of the false
annotation. A given enforcement mechanism might place a static well-formedness criterion
on the code it applies to. For instance, our\apt{or \citeauthor{DBLP:conf/sp/RoesslerD18}'s ? }
micro-policies expect that call and return
annotations occur on jump instructions. On the other hand, not all jumps are proper
calls or returns, only those preceded by appropriate sequences: in these cases the
bad annotation might appear well formed but the system should either respect the change
in permissions or, if doing so would compromise security, fail-stop rather than
execute the annotated instruction.\apt{hard to follow}

>>>>

%% Title information
\title{Stack Safety is a Security Property}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
%                                        %% can be repeated if necessary;
%                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
%                                        %% can be repeated if necessary;
%                                        %% contents suppressed with 'anonymous'

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  What is stack safety? The phrase has been used to describe the protections
  offered by a variety of mechanisms for guarding against
  attacks that corrupt the call stacks of running programs, including micro-policies
  implemented using metadata tagging mechanisms such as PIPE. But for specification
  these implementations largely appeal either to intuitive notions of the behavior
  of a stack, or to specific examples of bad behavior.
  %
  We propose a formal definition phrased in the technical terminology of language-based
  security---as a combination of an integrity property (a caller's stack frame is
  invariant over calls) and a confidentiality property (callees lack knowledge of the
  contexts of their calls). These combine with a straightforward well-bracketedness
  condition on the program's control flow that ensures that callers and callees remain
  meaningfully distinct.
  %
  While our formalism is informed by the PIPE implementation, the primary properties reflect
  general principles of stack safety. We define additional properties tailored for specific
  contexts: a stronger property amenable to efficient testing and a weaker one that is
  approximated by more efficient, lazy policies.

@@

But what {\em is} the stack safety that these techniques enforce? Precise formal
definitions are mostly lacking: rather, the notion is commonly defined in a
negative way, by enumerating the bad things that a given enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%\bcp{Does use of unitialized variables lead to stack safety attacks?\apt{sure, via buffer overflow}}
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
Indeed, we are only aware of one quite recent attempt to frame a positive definition, by
\citet{Skorstengaard+19}, which defines stack safety as the conjunction of two properties:
{\em local state encapsulation} and {\em well-bracketed control flow}.
These properties are formalized by defining a capability-based machine
that ``satisfies [stack safety] by construction.''
%
While this work represents an important step in the right direction, we believe
there is still room to craft an operational characterization
of stack safety. \apt{Perhaps our rhetoric should be more micro-policy-specific here?}
In other words, we seek formal criteria that can be applied
to a mechanism for protecting stack-based low-level programs to judge
whether or not it actually ``guarantees stack safety.'' We begin with a strong
local state encapsulation property that closely models the behavior of eager
micro-policies from \citet{DBLP:conf/sp/RoesslerD18}. This ``eager'' property
is too strong to characterize stack safety; reasonable enforcement mechanisms
exist that violate it. One such mechanism is a ``lazy'' tag policy, similar to
the one in \citet{DBLP:conf/sp/RoesslerD18}. We weaken local state encapsulation
to support this technique and argue that the new property, defined relative to an
arbitrary notion of observation. This property is still strong enough to prevent
the bad behaviors we associate with unsafe stack behavior, but weak enough that
other reasonable enforcement mechanisms can enforce it.

While our criteria are designed with PIPE micro-policies in mind, they are
agnostic about many implementation details and should be applicable to other
systems with minor adjustment.
%
Our key contribution is an operational model of local state encapsulation
using the technical framework of language-based security
\apt{this idea deserves more prominence}, with eager and lazy variants,
together with a simple characterization of well-bracketed control flow.

@@

How does the programmer rely on stack safety? The stack makes implicit promises about
the behavior of function calls. First, that when a call is finished, the caller can continue
from where it left off with its local state unchanged. And second, that the call's
behavior is determined only by a predictable interface that does not include the caller's
local state.

@@

Both expectations imply predictable control flow, termed {\em well-bracketed
control flow} (WBCF) by \citeauthor{Skorstengaard+19}. Calls always jump to valid entry
points and return to their call sites. Then the data portion of the first expectation is for
{\em integrity}: that a caller's local data is protected from modification by its callee.
This is vital for local reasoning about the behavior of the caller. The security implications
are significant: without integrity, a caller that branches on a local variable could be
manipulated by its callee into unexpected behavior.

Alongside integrity, we have {\em confidentiality}, in which a caller's data is protected
from leakage by its callee. As a reasoning principle, confidentiality tells us that to
predict the behavior of a call we need only consider a subset of the machine state.
And in security terms, we can guarantee the secrecy of local data that we haven't passed
even when calling a badly behaved function. Together, integrity and confidentiality
comprise \citeauthor{Skorstengaard+19}'s {\em local state encapsulation} (LSE) property.

@@

To phrase LSE and WBCF as security properties requires an explicit attacker model:
a clear understanding of who is being protected and from what. Most existing
stack protection mechanisms envisage scenarios where a callee is
influenced by external inputs to corrupt the stack in some way. To
avoid getting into the details of exactly how this happens when
formalizing our properties, we use an even stronger model: when we
state the protections that a function can rely on when it makes a
call, we assume that the callee can execute arbitrary instructions
until it returns. Thus we have no need to differentiate safe code
from adversarial - all functions, and indeed all activations of the same function,
are mutually distrusting.
%
For this model to make sense, we need to know when a call or a return is taking
place. We assume that calls and returns are annotated in the code, so that
we can track the point at which control transfers from the protected caller
to the attacking callee -- and when that callee in turn becomes a protected
caller in its own right.

@@

First we introduce this model for a single simple stack with no sharing between callers and
callee. Then we expand it to allow passing of stack data, granting a callee one time
access to the caller's data. We do not model more sophisticated sharing, because to do
so realistically would require a separate protection scheme, such as a capability
model. While this is quite feasible in a tag-based enforcement mechanism, it is properly
a separate policy outside the scope of stack safety. (A tag-based approach can easily
compose multiple such policies.)

Finally, we extend the model to a coroutine system featuring finitely-bounded stacks.
Though simplistic, this extension is a step toward concurrency and demonstrates that
the model is flexible enough to handle more sophisticated styles of control-flow.

After formally defining stack safety, we show that the definition is
enforceable using an existing tag-based enforcement
mechanism~\citep{DBLP:conf/sp/RoesslerD18}. We first consider
\citeauthor{DBLP:conf/sp/RoesslerD18}'s conservative Depth Isolation policy.
This policy testably enforces our {\em eager} properties, in which integrity is respected
at every step of a program, and each call respects integrity at each of its internal steps.
These properties are very strong, and our policies enforce them; this also makes them very
intensional, in that they do not distinguish internal behavior from external.
They also have the advantage that they are suitable for efficient testing, as they
fail immediately when a violation occurs.

The downside of the Depth Isolation policy is that it is rather expensive to implement.
In the same work, \citeauthor{DBLP:conf/sp/RoesslerD18} propose a different mechanism
that is more efficient by virtue of being lazy in its enforcement.
Instead of enforcing the integrity and confidentiality of the stack at the
exact moment when a callee returns, this policy signals violations when the
caller actually accesses data that the callee has written into its stack
frame. \ifaftersubmission\bcp{We could also hint at the lazy confidentiality
  property in the same way...}\fi
The lazy policy does not enforce our eager integrity property, and to our
knowledge there is no formal statement of the properties that it ought to enforce.
We propose {\em lazy} stack-safety properties, which capture the desired behavior of a
policy that detects integrity violations between when they occur and when they make
a visible impact. Intuitively this can be phrased as ``no visible reads after dangerous
writes.'' Unfortunately \citeauthor{DBLP:conf/sp/RoesslerD18}'s existing
``Lazy Tagging and Clearing'' policy does not quite enforce this property, as we show in
testing, but formalizing the ideal property allows us to see how it falls short.
We provide a variant of ``Lazy Tagging and Clearing'' that does enforce lazy integrity
at some performance cost.
%We prove (in Coq) that the eager property implies the lazy one.\apt{do we?
%  and is it so interesting anyhow?}
We argue that any system purporting to enforce stack safety ought to obey at least lazy
stack safety, making it a plausible extensional characterization.

@@

In this very strong attacker model, there is little obvious distinction between
the attacker and the defender. They are separated only by the control transfer
inherent in their calls and returns. Our stack-safety concepts therefore depend
critically on the notions of ``call'' and ``return,'' which are not fully explicit
in machine code (the same instruction opcode may be used in some places as part of
a call or return sequence and in other places for other purposes).
%
We assume we are given a machine language program annotated with the locations of
instructions representing calls, returns, and function entry points. Although this
information would typically come from a compiler together with the machine code,
we do not assume anything about their provenance. This means that attackers (or
foolish defenders) are under no obligation to have reasonable annotations.

@@

Alternatively, suppose an annotated instruction is not really an actual call or return.
What happens in this case depends on the enforcement mechanism and the nature of the false
annotation. A given enforcement mechanism might place a static well-formedness criterion
on the code it applies to. For instance, our\apt{or \citeauthor{DBLP:conf/sp/RoesslerD18}'s ? }
micro-policies expect that call and return
annotations occur on jump instructions. On the other hand, not all jumps are proper
calls or returns, only those preceded by appropriate sequences: in these cases the
bad annotation might appear well formed but the system should either respect the change
in permissions or, if doing so would compromise security, fail-stop rather than
execute the annotated instruction.\apt{hard to follow}
